Variation in use of East Asian Late Paleolithic weapons: A study of tip cross-sectional area of stemmed points from Korea

### 2023-02-17 analysis/paper/paper.qmd
Edit done by discussion

In order to explore spatial patterns and the relationship between the role of stemmed points and the environment, we located individual assemblages on maps that depict eco-regional zones. Our results show that stemmed points are only located in certain zones, Central Temperature Zone (CT) of the vegetation map or Central inland (CI) (@fig-gis-tcsa-sites panel A & D) of the eco-region map. We observed that there are statistically significant differences in the distributions of TCSA values between some zones (Vegetation zones: `r eco_lgm_test_pretty`, Ecoregion zones: `r for_lgm_test_pretty`). The Tukey's HSD results show that TCSA values from the South Temperature Zone (ST) zone are significantly different from the other two vegetation zones (@fig-gis-tcsa-sites panel C). Looking into this further, we see that there are only six artifacts from the ST zone, with only three of these having TSCA values above 175. The other three have TCSA values lower than 100, similar to the other zones (@fig-gis-tcsa-sites panel B). Given the small number of artifacts with extreme TCSA values in the ST zone, we hesitate to conclude this is an archaeologically significant pattern of different TCSA values. The Tukey HSD pairwise differences show no pairs of eco-regions with significant differences in TCSA values, likely due to the small number of points in several zones (@fig-gis-tcsa-sites panel F). Overall, our results indicate no clear patterns in the distribution of TCSA values across the vegetation and eco-regional zones.

@@

To investigate the role that stemmed points played in the technological transition during the Korean Late Paleolithic, we computed the TCSA metric to infer likely uses by comparing the results to the TCSA range from other archaeological and ethnographic cases (@tbl-tcsa-ranges). Our main questions were: What were the best-fit ballistic probabilities for the stemmed points if they were hafted as weapon tips? How diverse were their likely uses? What are the temporal and spatial patterns of stemmed point uses?

Our results show that javelin tips and stabbing spear tips are the most probable ballistic uses for stemmed points. There are a few stemmed points in our sample that may have been used as dart tips and un-poisoned arrow tips. None have TCSA values in the range of poison arrow tips (@fig-tcsa-all-sp). @lee2019were computed TCSA for 10 stemmed points from Jingeuneul site, which we were not able to include in this research due to the unavailability of their data, and concluded that their TCSA values fall within the range of North American dart tips and arrowheads. In general, we found a wider range of TCSA in our sample, indicating that stemmed points might play diverse roles in foraging toolkits.

TCSA can be impacted by other factors such as raw materials and portability. @eren2022plains explain that smaller TCSA could be the result of pursuing production economy and transport efficiency. We explored the relationship between TCSA and raw material types and weight in order to examine these factors. We observed no clear pattern for raw materials (@fig-tcsa-raw-materials) but weight (@fig-artifact-size-cluster), though weight is correlated with raw materials (@fig-tcsa-size). We speculate this might be due to the higher availability of raw materials such as shale on hornfels on the landscape. Therefore we could assume that raw materials also associated with TCSA value. We found that different clusters of weight are matched with different types of weapon tips (@fig-artifact-size-cluster).

Our results shows that TCSA values vary between assemblages with few discernible temporal patterns in function (@fig-tcsa-radio-carbon panel A). Stemmed points were used most frequently before the LGM (@fig-tcsa-radio-carbon panel B). While we predicted the greatest variation in TCSA values during the LGM, our results show that variability decreased during the LGM, in comparison to the previous period. This could be explained by the overall decline in stemmed points usage.

We found that stemmed points are most frequently located in a small number of eco-regional zones such as the Central Temperature zone with few clears pattern in TCSA function across the zones (@fig-gis-tcsa-sites). As predicted, TCSA values in inland areas, such as vegetation zone CT, show higher variability. Based on these results, we assume that stemmed points performed a wider range of tasks in low-productivity patches. Both temporal and spatial analyses show that stemmed points were likely more adapted for particular environments, rather than specific climate contexts. @prates2022changes claimed that fishtail projectile points in South America was used for megafauna hunting and contributed to their extinction by proving strong relationship between the temporal density and spatial distribution of megafauna and the projectile points. We similarly found that stemmed points are more densly distributed in certain environments in South Korea, perhaps also relating to the distribution of fauna.

Overall, our temporal and spatial patterns of TCSA show a wide range of TCSA values throughout the Late Paleolithic period and between eco-regional zones. We confirmed the widest TCSA range in a single site, Suyanggae (@fig-tcsa-per-sites). As a consequence, our results are consistent with our second scenario in which stemmed points are described as multifunctional tools, suggesting that people created stemmed points in response to unexpected or varying circumstances in their habitats, similar to how @eren2022plains explain Clovis points.

>>>>

```
In order to explore spatial patterns and the relationship between the role of stemmed points and the environment, we located individual assemblages on maps that depict eco-regional zones. Our results show that stemmed points are only located in certain zones, Central Temperature Zone (CT) of the vegetation map or Central inland (CI) (@fig-gis-tcsa-sites panel A & D) of the eco-region map. We observed that there are statistically significant differences in the distribution of TCSA values between some zones (Vegetation zones: `r eco_lgm_test_pretty`, Ecoregion zones: `r for_lgm_test_pretty`). The Tukey's HSD results show that TCSA values from the South Temperature Zone (ST) zone are significantly different from the other two vegetation zones (@fig-gis-tcsa-sites panel C). Looking into this further, we see that there are only six artifacts from the ST zone, with only three of these having TSCA values above 175. The other three have TCSA values lower than 100, similar to the other zones (@fig-gis-tcsa-sites panel B). Given the small number of artifacts with extreme TCSA values in the ST zone, we hesitate to conclude this is an archaeologically significant pattern of different TCSA values. The Tukey HSD pairwise differences show no pairs of eco-regions with significant differences in TCSA values, likely due to the small number of points in several zones (@fig-gis-tcsa-sites panel F). Overall, our results indicate no clear patterns in the distribution of TCSA values across vegetation and eco-regional zones.

@@

By comparing the results with the TCSA range from other archaeological and ethnographic cases (@tbl-tcsa-ranges), we were able to determine what role stemmed points played in the technological transition during the Korean Late Paleolithic. Our main questions were: What were the best-fit ballistic probabilities for stemmed points if they were hafted as weapon tips? How diverse were their likely uses? What are the temporal and spatial patterns of stemmed point uses?

Our results indicate that javelin tips and stabbing spear tips are the most probable ballistic uses for stemmed points. There are a few stemmed points in our sample that may have been used as dart tips and un-poisoned arrow tips. None have TCSA values in the range of poison arrow tips (@fig-tcsa-all-sp). In their study, @lee2019 calculated TCSA values for ten stemmed points from Jingeuneul site, which were not included in the research due to data availability issues, and concluded that these values are within the range of North American dart tips and arrowheads. In general, we found a wide range of TCSA in our sample, suggesting that stemmed points might play diverse roles in foraging toolkits.

TCSA can be impacted by other factors such as raw materials and portability. @eren2022plains explain that smaller TCSA could be the result of pursuing production economy and transport efficiency. We explored the relationship between TCSA and raw material types and weight in order to examine these factors. We observed no clear pattern for raw materials (@fig-tcsa-raw-materials) but weight (@fig-artifact-size-cluster), though weight is correlated with raw materials (@fig-tcsa-size). We speculate that this might be due to the higher availability of raw materials, such as shale and hornfels on the landscape. Therefore, we could assume that raw materials also associated with TCSA value. We found that different clusters of weight are matched with different types of weapon tips (@fig-artifact-size-cluster).

Our results show that TCSA values vary between assemblages with few discernible temporal patterns in function (@fig-tcsa-radio-carbon panel A). Stemmed points were used most frequently before the LGM (@fig-tcsa-radio-carbon panel B). While we predicted the greatest variation in TCSA values during the LGM, our results show that variability decreased during the LGM, in comparison to the previous period. This could be explained by the overall decline in stemmed points usage.

We found that stemmed points are primarily located in a small number of eco-regional zones such as the Central Temperature zone with few clear patterns in TCSA function across the zones (@fig-gis-tcsa-sites). As predicted, TCSA values in inland areas, such as vegetation zone CT, show higher variability. These results suggest that stemmed points performed a wider range of tasks in low-productivity patches. Both temporal and spatial analyses show that stemmed points were more likely to have adapted to particular environments, rather than specific climate contexts. @prates2022changes claimed that fishtail projectile points in South America were used to hunt megafauna and contributed to their extinction by demonstrating a strong correlation between the spatial and temporal distribution of megafauna and the projectile points. We similarly found that stemmed points are more densely distributed in certain environments in South Korea, perhaps also related to faunal distribution.

Overall, our results show a wide range of TCSA values throughout the Late Paleolithic period and between eco-regional zones. We confirmed the widest TCSA range at a single site, Suyanggae (@fig-tcsa-per-sites). Our foundings are therefore consistent with our second scenario, in which stemmed points are described as multifunctional tools, suggesting that people created stemmed points as a response to unexpected or varying circumstances in their habitats, similar to how @eren2022plains explain Clovis points.
